Clinical prediction using similarity cohort--based models: A systematic scoping review

### 2025-03-07 docs/Localized-Modeling-BMCAI/localized-modeling-dai-revision.Rmd
fix typos & update latex source

We do observe one clear pattern: All of the proposed methods that most evidently outperformed their comparators---Kohonen + CBR [@Wyns2004], TWNFI [@Song2006, @Kasabov2010], gravitational search algorithms (GSA) [@Liang2015], CBR + rules (+ DML) [@Nicolas2014], Gaussian process regression [@Zhang2018], JITL-ELM [@Ma2020]---are hybrids of \hl{SCBM} (in some cases CBR) with other techniques, often DML. Though @CampilloGimenez2013 and @Ng2015 report non-superior performance by such hybrids, the pattern suggests the importance of the similarity measure to the retrieval step.
Meanwhile, when proposed approaches targeted cohort demarcation or choice of predictive model---statistical CBR [@Park2006]; individualized \hl{GLM, DT, and RF} [@Lee2015; @Lee2017]---they did not consistently outperform comparators.

@@

These programs used varying terminology for common concepts, and no common term is in use for what we here refer to as \hl{similarity cohort--based modeling}; authors described their approaches as "targeted prognosis" [@Mariuzzi1997], "transductive inference" [@Song2006], "personalized decision support" [@Lee2015], "personalized (predictive) models" [@Liang2015, in contrast to "local models"; @Ng2015; @Wang2019; @Ma2020; @Liu2022; @Doborjeh2022], and "precision cohort" analysis [@Wang2019; @Tang2021; @Ng2021]. We find uses of "targeted", "personalized", and "precision" generic and imprecise, while transductive inference is an established term for a broader set of methods in ML.
\hl{While "local" naturally contrasts with "global", the overloaded term "localized models" stands to cause more confusion than it resolves, so we use "similarity cohort--based models" to refer to this specific approach.}

@@

\hl{The structure and terms of our search strings were inspired in part by previous reviews adjacent to our topic of interest }[@Sharafoddini2017; @Parimbelli2018]\hl{. We organized the PubMed search in disjunctive normal form (an OR of ANDs). Following the solidification of an outline, we expanded each term to include synonyms that are similar enough to be applicable to our search. We then evaluated the expanded search string using the PubMed Advanced Search platform. We initially included each term and their synonyms separately to evaluate what resulted. We pruned several terms that returned no results ("phrases not found") or to reduce the number of results. At the conclusions of this process for each individual term, we combined the search terms using disjunctive normal form.}

@@

\hl{As one reviewer pointed out for many if not most models used in ML a predictor may play a different and more or less important role for some cases than others, as quantified by importance measures or model-agnostic explanations }[@Biecek2021; @Molnar2023]\hl{. In this way, such models can also be viewed as "localized". We recognize, for example, that the use of global and local explanations is standard. As another reviewer pointed out, the terminology of local and global is also used in the unrelated context of federated ML to distinguish models trained using data stored in a single storage device (local) from their aggregations (global) [@Moshawrab2023; @Brauneck2023]. The prospects for federated architecture to more effectively implement similarity cohort--based models are intriguing, and would require some reconciliation of terms.
A third reviewer pointed out that "local" may describe models fitted to data collected at a specific site or using features extracted from a specific patch of an image. Taken together, these issues convinced us to drop the shorthand "localized model" and adopt the more specific term "similarity cohort--based model".}

>>>>

We do observe one clear pattern: All of the proposed methods that most evidently outperformed their comparators---Kohonen + CBR [@Wyns2004], TWNFI [@Song2006; @Kasabov2010], gravitational search algorithms (GSA) [@Liang2015], CBR + rules (+ DML) [@Nicolas2014], Gaussian process regression [@Zhang2018], JITL-ELM [@Ma2020]---are hybrids of \hl{SCBM} (in some cases CBR) with other techniques, often DML. Though @CampilloGimenez2013 and @Ng2015 report non-superior performance by such hybrids, the pattern suggests the importance of the similarity measure to the retrieval step.
Meanwhile, when proposed approaches targeted cohort demarcation or choice of predictive model---statistical CBR [@Park2006]; individualized \hl{GLM, DT, and RF} [@Lee2015; @Lee2017]---they did not consistently outperform comparators.

@@

These programs used varying terminology for common concepts, and no common term is in use for what we here refer to as \hl{similarity cohort--based modeling}; authors described their approaches as "targeted prognosis" [@Mariuzzi1997], "transductive inference" [@Song2006], "personalized decision support" [@Lee2015], "personalized (predictive) models" [@Liang2015, in contrast to "local models"; @Ng2015; @Wang2019; @Ma2020; @Liu2022; @Doborjeh2022], and "precision cohort" analysis [@Wang2019; @Tang2021; @Ng2021]. We find uses of "targeted", "personalized", and "precision" generic and imprecise, while transductive inference is an established term for a broader set of methods in ML.
\hl{While ``local'' naturally contrasts with ``global'', the overloaded term ``localized models'' stands to cause more confusion than it resolves, so we use ``similarity cohort--based models'' to refer to this specific approach.}

@@

\hl{The structure and terms of our search strings were inspired in part by previous reviews adjacent to our topic of interest }[@Sharafoddini2017; @Parimbelli2018]\hl{. We organized the PubMed search in disjunctive normal form (an OR of ANDs). Following the solidification of an outline, we expanded each term to include synonyms that are similar enough to be applicable to our search. We then evaluated the expanded search string using the PubMed Advanced Search platform. We initially included each term and their synonyms separately to evaluate what resulted. We pruned several terms that returned no results (``phrases not found'') or to reduce the number of results. At the conclusions of this process for each individual term, we combined the search terms using disjunctive normal form.}

@@

\hl{As one reviewer pointed out for many if not most models used in ML a predictor may play a different and more or less important role for some cases than others, as quantified by importance measures or model-agnostic explanations }[@Biecek2021; @Molnar2023]\hl{. In this way, such models can also be viewed as ``localized''. We recognize, for example, that the use of global and local explanations is standard. As another reviewer pointed out, the terminology of local and global is also used in the unrelated context of federated ML to distinguish models trained using data stored in a single storage device (local) from their aggregations (global) }[@Moshawrab2023; @Brauneck2023]\hl{. The prospects for federated architecture to more effectively implement similarity cohort--based models are intriguing, and would require some reconciliation of terms.
A third reviewer pointed out that ``local'' may describe models fitted to data collected at a specific site or using features extracted from a specific patch of an image. Taken together, these issues convinced us to drop the shorthand ``localized model'' and adopt the more specific term ``similarity cohort--based model''.}
